Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-04-09 ThesisJP/Background/background.tex
Started working on comments from CI.

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate until the required cooling finally became to much in 2004.  From that point, chip manufactures had to pursue a different direction  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

@@

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is designed to be the main control unit of a personal computer, and is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Although they started out as very different processors, modern CPUs are now including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these features are and why there is still a big difference in peak performance between a CPU and a GPU.

@@

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation. Neither gamma rays nor high intensity magnetic fields are involved, just sound pressure below well-controlled limits. It is also far less costly than other medical imaging modalities and offers real-time inspection of the human body.  Ultrasound imaging is well known for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging. An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}. Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}. Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle. The tissue separating the left and right ventricle is known as the inter-ventricular septum. The left side is responsible of pumping blood out to the whole body through the aortic valve. It is therefore the strongest and hardest working part of the heart. When blood flows out of the aorta it is important that the mitral valve, which separates the left atrium from the left ventricle, is closed and does not leak. The left side, and especially the left ventricle and the mitral valve, therefore receives the most attention during cardiac ultrasound examinations. When visualizing 3D ultrasound images in \textbf{Paper IV} we will have a strong focus on visibility of cardiac tissue, thus note for later that the endocardium refers to the innermost layer of the cardiac muscle.

@@

To generate the ultrasonic frequencies used for imaging, a piezoelectric ceramic is used. A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality. An example of a probe for medical ultrasound imaging is shown in Fig.\,\ref{fig:us_grid}. %For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size (the \textit{aperture}) of the array grows. The angular width of this beam, and also the angular resolution of the array, is approximately given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution. The geometry of cardiac ultrasound imaging is depicted in Fig.\,\ref{fig:us_grid}. A line in an ultrasound image is generated by firing an ultrasound pulse or beam into the body and sample the received echo. We will refer to this amplitude trace as an \textit{image line}.

\subsection{Beamforming}
If time delays are applied on the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$ radians. As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ radians wide. This is known as a 2D B-mode image (B for brightness). If image lines are acquired for multiple B-mode images which are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$-plane. In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

@@

The standard form of beamforming used on receive in ultrasound imaging is known as \textit{delay-and-sum} (DAS) beamforming: 
\begin{align}\label{eq:das_background}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\Delta_m[n]$.  From now on we will omit the dependency on $\Delta_m[n]$ and $x_m$ will refer to delayed element data. A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das_background}. If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse. Transmit delays therefore do not vary within an image line.

\subsection{Sampling and processing complexity}
When generating a B-mode image we have to insure that the spatial and temporal sampling are sufficient. We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system which both emit and receive energy is governed by the sum of the transmit ($D_{tx}$\nomi{tx}{Transmit}) and receive ($D_{rx}$\nomi{rx}{Receive}) aperture. The lateral resolution of what is known as an active imaging system is found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}image lines distributed across the imaging sector. If \nom{$r$}{Imaging range.} is the maximum range in our image, as shown in Fig.\,\ref{fig:us_grid}, we can acquire
\begin{align}
f_{\text{FPS}} = \frac{c}{2rN_\theta}
\end{align} 
images or frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short pulse consists of many frequency components. A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth. Medical ultrasound systems typically has a bandwidth between 33 \% and 100 \% of the pulse frequency. This corresponds to transmitting a pulse which is between 1 and 3 oscillations when entering tissue. In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{in-phase quadrature (\nom{IQ}{In-phase quadrature (ultrasound data format).}) demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one image line is after this process given by:
\begin{align}
N = \frac{2r}{c}B_wf_c,
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}, where $f_c$ is the center frequency of the transmitted pulse. Note that after this process the ultrasound data will be complex values. The total number of samples in an B-mode image from a phased array is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of complex element data every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c.
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. Hence, far below the peak processing throughput of modern GPUs and CPUs. The main challange with DAS beamforming is the amount of data that has to be moved. In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5. Newer versions of PCI Express which supports even higher transfer rates have also recently been introduced.

@@

\subsection{GPUs in medical ultrasound imaging}
Ultrasound scanners generate an incredible amount of data. A modern system capable of 3D imaging will generate several tera bytes of data inside the probe handle during a clinical scan. Fortunately the data rate is reduced to the level of 2D imaging before the data reaches the scanner. The data rate of a 2D scan is "only" in order of giga bytes per second. To handle this amount of data, ultrasound imaging systems have, for the last twenty years, relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline \cite{Thomenius}. However, as the computational power of generic processors have increased, ASICs have gradually been replaced by software \cite{Guracar2013}. One of  the last remaining ASICs in ultrasound scanners is used for beamforming \cite{Thomenius}. As seen from (\ref{eq:das_background}), beamforming involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe. The input data rate is therefore very high. Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}. This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}. We therfore can conclude that beamforming is finally moving into software too. The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}. Software beamforming opens up for new possibilities for advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which is introduce in the next section. The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.

@@

To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed in the literature \cite{Asl2012, Kim}. Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}. In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion. In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of the element data: 
\begin{align}
\vec{x}_{\text{BS},l} &= \mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}.
\end{align}
The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$. When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that it is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.

>>>>

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate until the required cooling finally became to much in 2004.  From that point, CPU manufactures had to pursue a different direction  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

@@

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is designed to be the main control unit of a personal computer, and is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Although they started out as very different processors, modern CPUs are now including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these features are and why there is still a big difference in theoretical peak performance between a CPU and a GPU.

@@

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation of the human body. Neither gamma rays nor high intensity magnetic fields are involved, just sound pressure below well-controlled limits. It is also far less costly than other medical imaging modalities and offers real-time imaging. Ultrasound imaging is well known for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging. An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}. Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}. Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle. The tissue separating the left and right ventricle is known as the inter-ventricular septum. The left side is responsible of pumping blood out to the whole body through the aortic valve. It is therefore the strongest and hardest working part of the heart. When blood flows out of the aorta, it is important that the mitral valve, which separates the left atrium from the left ventricle, is closed and does not leak. The left side, and especially the left ventricle and the mitral valve, therefore receives the most attention during cardiac ultrasound examinations. When visualizing 3D ultrasound images in \textbf{Paper IV} we will have a strong focus on visibility of cardiac tissue, thus note for later that the endocardium refers to the innermost layer of the cardiac muscle.

@@

To generate the ultrasonic frequencies used for imaging, a piezoelectric ceramic is used. A piezoelectric material has a special property where mechanical vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. A set of piezoelectric elements organized in an array is referred to as an ultrasound \textit{probe}, and special probes are typically designed for different medical imaging modalities. An example of a probe for medical ultrasound imaging is shown in Fig.\,\ref{fig:us_grid}. %For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array, the transmitted pulse will get shaped into a beam\footnote{A beam refers to the accumulated magnitude across time of a pulse traveling in space.} of ultrasound by diffraction as the width (the \textit{aperture}) of the array grows. The angular width of this beam, and also the angular resolution in radians of the array, is approximately given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} are the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width. With an average speed of sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree. The axial resolution is inversely proportional to the pulse length, and is typically better than the lateral resolution. The geometry of cardiac ultrasound imaging is depicted in Fig.\,\ref{fig:us_grid}. A line in an ultrasound image is generated by firing an ultrasound pulse or beam into the body and sample the received echo. We will refer to this amplitude trace as an \textit{image line}.

\subsection{Beamforming}
If time delays are applied to the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$ radians. As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ radians wide. This is known as a 2D B-mode image (B for brightness). If image lines are acquired for multiple B-mode images that are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$ planes. In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

@@

The standard form of beamforming used in ultrasound imaging is known as \textit{delay-and-sum} (DAS): 
\begin{align}\label{eq:das_background}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\Delta_m[n]$.  From now on we will omit the dependency on $\Delta_m[n]$ and $x_m$ will refer to delayed element data. A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das_background}. If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. When transmitting the pipeline depicted in Fig.\,\ref{fig:das_background}, without the sum operator, is reversed and the negated set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse. Transmit delays therefore do not vary within an image line.

\subsection{Sampling and processing complexity}
When generating a B-mode image, we have to insure that the spatial and temporal sampling are sufficient. We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system that both emits and receives energy is governed by the sum of the transmit ($D_{tx}$\nomi{tx}{Transmit}) and receive ($D_{rx}$\nomi{rx}{Receive}) aperture. The lateral resolution of what is known as an active imaging system is found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}image lines distributed across the imaging sector. If \nom{$r$}{Imaging range.} is the maximum range in our image, as shown in Fig.\,\ref{fig:us_grid}, we can acquire
\begin{align}
f_{\text{FPS}} = \frac{c}{2rN_\theta}
\end{align} 
images or frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short pulse consists of many frequency components. A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth. Medical ultrasound systems typically has a bandwidth between 33 \% and 100 \% of the pulse frequency. This corresponds to transmitting a pulse which is between 1 and 3 oscillations when entering tissue. In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{in-phase quadrature (\nom{IQ}{In-phase quadrature (ultrasound data format).}) demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one image line is after this process given by:
\begin{align}
N = \frac{2r}{c}B_wf_c,
\end{align}
\nomi{$N$}{Number of samples in range (or time)}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}, where $f_c$ is the center frequency of the transmitted pulse. Note that after this process the ultrasound data will be complex values. The total number of samples in an B-mode image from a phased array is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ complex data vectors of length $M$ every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c.
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. This is far below the peak processing throughput of modern GPUs and CPUs. The main challange with DAS beamforming is the amount of data that has to be moved. In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5. Newer versions of PCI Express, which supports even higher transfer rates, have also recently been introduced.

@@

\subsection{GPUs in medical ultrasound imaging}
Ultrasound scanners generate an incredible amount of data. A modern system capable of 3D imaging will generate several tera-bytes of data inside the probe handle during a clinical scan. Fortunately the data rate is reduced to the level of 2D imaging before the data reaches the scanner. The data rate of a 2D scan is "only" in order of giga-bytes per second. To handle this amount of data, ultrasound imaging systems have, for the last twenty years, relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline \cite{Thomenius}. However, as the computational power of generic processors have increased, the expensive ASICs have gradually been replaced by software \cite{Guracar2013}. One of  the last remaining ASICs in ultrasound scanners is used for beamforming \cite{Thomenius}. As seen from (\ref{eq:das_background}), beamforming involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe. The input data rate is therefore very high. Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}. This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}. We can therefore conclude that beamforming is finally moving into software too. The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}. Software beamforming opens up for cost-efficient and rapid prototyping of advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}) and introduced in the next section. The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.

@@

To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed in the literature \cite{Asl2012, Kim}. Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}. In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion. In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of the element data: 
\begin{align}
\vec{x}_{\text{BS},l} &= \mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}.
\end{align}
The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$. When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that it is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace (BS-Capon)} is the number of selected beams.
